---
title: "Why Have House Prices in Turkey Gone Up in the Last Ten Years?"
author:
  - name: "Ömer Furkan Demirci"
  - name: "Fatih Alper Vural"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    fig-width: 9
    fig-height: 5
    fig-cap-location: bottom
    df-print: paged
  pdf:
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    margin-left: 2cm
    margin-right: 2cm
    margin-top: 2cm
    margin-bottom: 2cm
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Abstract {.unnumbered}

Housing affordability has emerged as one of the most pressing socioeconomic challenges in Türkiye over the past decade. This study investigates the key drivers of the sharp and sustained rise in housing prices between 2014 and 2024, drawing on four official monthly time-series datasets obtained from the Central Bank of the Republic of Türkiye (TCMB/EVDS): the House Price Index (HPI), housing sales statistics, USD/EUR exchange rates, and credit card sectoral expenditure data. Through exploratory analysis, correlation testing, regional comparison, and multiple linear regression modelling, we find that exchange rate depreciation and broad consumer spending inflation are the dominant macroeconomic forces behind rising house prices, while sales volumes display an inverse relationship with price levels — consistent with demand compression at high price levels. Regional heterogeneity is substantial: İstanbul and western metropolitan areas lead price growth, whereas interior Anatolian regions lag considerably. Our model explains approximately 95% of the variance in the national HPI, suggesting that macroeconomic indicators are strong predictors of housing price trajectories in Türkiye.

------------------------------------------------------------------------

# Project Overview and Scope

We want to look into the following question for this project:\
\
**Why have house prices in Turkey gone up in the last ten years?**\
\
To answer this question, we made a dataset using four monthly time-series datasets from the Central Bank of the Republic of Türkiye's (TCMB) Electronic Data Delivery System (EVDS). These datasets have:\

-   Housing Price Index

-   Sales of Houses

-   Exchange Rates (USD/TRY and EUR/TRY)

-   Credit Card Sectoral Expenditures\
    \

We chose this topic because the cost of housing has become one of the most important social and economic issues in Türkiye. It is important to know what caused this rise in order to make good decisions about policy and the economy. This problem is also a good fit for a data analytics approach because it needs to combine multiple datasets, do preprocessing, and look at how things change over time and in different areas.

We want to find the main things that cause housing prices to go up. Specifically, we want to look at:

-   How exchange rates affect the price of homes

-   The link between changes in housing prices and demand (sales)

-   The function of consumption behavior (credit card expenditure) as an indicator of demand conditions.

-   Differences in how housing works in different regions

# Data

You can see Our data in R Data format below.

[Project Data](https://github.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/blob/e273d93492dc0a2ea56f85a472a398fe4909c2d7/Data/Project_Data.RData)

```{r}
url <- "https://raw.githubusercontent.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/main/Data/Project_Data.RData"
download.file(url, destfile = "Project_Data.RData", mode = "wb")
load("Project_Data.RData")

ls()
```

## Data Source

The data used in this project were obtained from the **Electronic Data Delivery System (EVDS)** of the Central Bank of the Republic of Türkiye (TCMB). EVDS provides official, high-frequency economic data and is widely used for academic and professional analysis.

The project is based on four datasets:

-   **House Price Index (HPI)**\
    Monthly housing price index at national and regional levels

-   **House Sales Statistics**\
    Monthly housing sales data at the provincial level

-   **Exchange Rates (USD, EUR)**\
    Monthly exchange rate data

-   **Credit Card Sectoral Expenditures**\
    Monthly sector-based credit card spending data

## General Information About Data

The processed dataset consists of three main analytical tables:

### Exchange Rate Table

-   **Description**:\
    This table contains the columns Year, Month, USD Conversion Rate, and EUR Conversion Rate. Represents monthly exchange rate movements. These variables are used as indicators of macroeconomic conditions and cost-side pressures that may affect housing prices.

-   In general, the exchange rate variables show a strong upward trend over time rather than a stable distribution.

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(scales)
    usd_eur <- Conversion_Rates %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(usd_eur, aes(x = Date)) +
      geom_line(aes(y = USD, color = "USD")) +
      geom_line(aes(y = EUR, color = "EUR")) +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      scale_y_continuous(labels = label_number(accuracy = 0.1)) +
      labs(
        title = "Exchange Rates Over Time",
        x = NULL,
        y = "TRY per Foreign Currency",
        color = "Currency"
      ) +
      theme_minimal()
    ```

### Credit Card Expenditure Table

-   **Description**:

-   This table contains Year, Month, Expense Type, and Quantity.\

-   Expense Type identifies the spending category, and Quantity gives the corresponding spending amount for that month.

-   This table contains sectoral spending amounts derived from credit card transactions.

-   The values in this table are expected to differ substantially across categories. Some categories such as retail, food, fuel, and e-commerce are likely to have much larger values than others.

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(scales)
    expense_quantity <- Expense_Quantity %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(
      expense_quantity %>%
        filter(Expense_Type %in% c("E-Alışveriş", "Çeşitli Gıda", "Giyim ve Aksesuar", "Benzin ve Yakıt İstasyonları")),
      aes(x = Date, y = Quantity / 1e3, color = Expense_Type)
    ) +
      geom_line() +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      scale_y_continuous(labels = label_comma(accuracy = 1)) +
      labs(
        title = "Credit Card Expenditures for Selected Categories",
        x = NULL,
        y = "Spending (Billion TRY)",
        color = "Expense Type"
      ) +
      theme_minimal() +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8))
    ```

### Housing Panel Dataset

-   **Description**:\
    This is the main table for the project and contains Year, Month, Region, House Price Index, and Total Sales.

    Region represents the housing price index regions as in NUTS2 level, House Price Index shows the monthly house price index for that region, and Total Sales gives the aggregated monthly house sales value for the same region.

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(scales)

    konut_index_sales <- Konut_Index_Sales %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(
      konut_index_sales %>% filter(Region == "Türkiye"),
      aes(x = Date, y = Konut_Endeksi)
    ) +
      geom_line() +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      scale_y_continuous(labels = label_comma(accuracy = 1)) +
      labs(
        title = "House Price Index in Türkiye Over Time",
        x = NULL,
        y = "House Price Index"
      ) +
      theme_minimal()
    ```

```{r}
library(dplyr)
library(ggplot2)
library(scales)

ggplot(
  Konut_Index_Sales %>%
    mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month))) %>%
    filter(Region %in% c("Türkiye", "İstanbul", "İzmir", "Ankara", "Adana, Mersin")),
  aes(x = Date, y = Total_Sales, color = Region)
) +
  geom_line() +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Total House Sales for Selected Regions",
    x = NULL,
    y = "Units Sold",
    color = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Potential Analyses

-   time-series plots of house prices, exchange rates, and sales,
-   regional comparisons of house price growth,
-   correlations between exchange rates and the house price index,
-   relationships between sales volume and housing prices,
-   category-based spending trends over time,

## Reason of Choice

Housing prices are one of the most important economic and social issues in Türkiye. The rapid increase in housing prices has significant implications for affordability, investment decisions, and economic stability.

All datasets are:

-   Consistent in frequency (monthly)

-   Long enough for trend analysis

-   Rich enough for regional comparison

    This makes it easy for analyzing.

## Preprocessing

The raw EVDS datasets were provided in Excel format and contained different structures (wide vs. long) and aggregation levels (province vs. region). We performed preprocessing in Python using pandas to standardize date formats, reshape wide datasets into long format, map categorical codes to meaningful labels, and align all datasets at a consistent monthly and regional level. In particular, house sales data were converted from province-level columns to regional aggregates compatible with the House Price Index. The final result consists of three clean analytical tables ready for further analysis.

Another key issue was that EVDS datasets use **coded column names** (e.g., `KT2`, `T58`, `TR62`) instead of descriptive labels. To make the data interpretable and consistent across sources, we created mapping dictionaries in Python to translate these codes into meaningful categories, provinces, and regions.

Here you can find our python code.\
\

``` py
# We explained the data and the output tables to AI. It gave us this code's first version. Then we fixed it according to our needs
import pandas as pd
from pathlib import Path

downloads = Path.home() / "Downloads"

# Files
kurlar_file = downloads / "Kurlar.xlsx"
kredi_file = downloads / "Kredi Harcama.xlsx"
konut_satis_file = downloads / "Konut Satış.xlsx"
konut_fiyat_file = downloads / "Konut Fiyat Endexi.xlsx"

output_file = downloads / "evds_preprocessed_tables_fixed.xlsx"

# -----------------------------
# Helper
# -----------------------------
def split_year_month(df):
    df["Tarih"] = pd.to_datetime(df["Tarih"], format="%Y-%m")
    df["Year"] = df["Tarih"].dt.year
    df["Month"] = df["Tarih"].dt.month
    return df

# -----------------------------
# Read
# -----------------------------
kurlar = split_year_month(pd.read_excel(kurlar_file))
kredi = split_year_month(pd.read_excel(kredi_file))
satis = split_year_month(pd.read_excel(konut_satis_file))
kfe = split_year_month(pd.read_excel(konut_fiyat_file))

# -----------------------------
# TABLE 1: Exchange Rates
# -----------------------------
table_kurlar = (
    kurlar[["Year", "Month", "TP_DK_USD_S_YTL", "TP_DK_EUR_S_YTL"]]
    .rename(columns={
        "TP_DK_USD_S_YTL": "USD",
        "TP_DK_EUR_S_YTL": "EUR"
    })
)

# -----------------------------
# TABLE 2: Credit Expenditures
# -----------------------------
expense_type_map = {
    "TP_KKHARTUT_KT2": "Car Rental",
    "TP_KKHARTUT_KT3": "Car Sales/Service",
    "TP_KKHARTUT_KT4": "Fuel",
    "TP_KKHARTUT_KT5": "Food",
    "TP_KKHARTUT_KT6": "Direct Marketing",
    "TP_KKHARTUT_KT7": "Education",
    "TP_KKHARTUT_KT8": "Electronics",
    "TP_KKHARTUT_KT9": "Clothing",
    "TP_KKHARTUT_KT10": "Airlines",
    "TP_KKHARTUT_KT11": "Services",
    "TP_KKHARTUT_KT12": "Accommodation",
    "TP_KKHARTUT_KT13": "Social",
    "TP_KKHARTUT_KT14": "Entertainment",
    "TP_KKHARTUT_KT15": "Jewelry",
    "TP_KKHARTUT_KT16": "Retail",
    "TP_KKHARTUT_KT17": "Furniture",
    "TP_KKHARTUT_KT18": "Construction",
    "TP_KKHARTUT_KT19": "Health",
    "TP_KKHARTUT_KT20": "Travel",
    "TP_KKHARTUT_KT21": "Insurance",
    "TP_KKHARTUT_KT22": "Telecom",
    "TP_KKHARTUT_KT23": "Hardware",
    "TP_KKHARTUT_KT24": "Food Service",
    "TP_KKHARTUT_KT25": "Misc",
    "TP_KKHARTUT_KT26": "Other",
    "TP_KKHARTUT_KT50": "E-Commerce"
}

kredi_cols = [c for c in expense_type_map if c in kredi.columns]

table_kredi = (
    kredi.melt(
        id_vars=["Year", "Month"],
        value_vars=kredi_cols,
        var_name="Code",
        value_name="Quantity"
    )
    .assign(Expense_Type=lambda x: x["Code"].map(expense_type_map))
    [["Year", "Month", "Expense_Type", "Quantity"]]
)

# -----------------------------
# TABLE 3: Housing Panel
# -----------------------------
provinces = [
    "Adana","Adıyaman","Afyonkarahisar","Ağrı","Aksaray","Amasya","Ankara",
    "Antalya","Ardahan","Artvin","Aydın","Balıkesir","Bartın","Batman",
    "Bayburt","Bilecik","Bingöl","Bitlis","Bolu","Burdur","Bursa","Çanakkale",
    "Çankırı","Çorum","Denizli","Diyarbakır","Düzce","Edirne","Elazığ",
    "Erzincan","Erzurum","Eskişehir","Gaziantep","Giresun","Gümüşhane",
    "Hakkâri","Hatay","Iğdır","Isparta","İstanbul","İzmir","Kahramanmaraş",
    "Karabük","Karaman","Kars","Kastamonu","Kayseri","Kırıkkale","Kırklareli",
    "Kırşehir","Kilis","Kocaeli","Konya","Kütahya","Malatya","Manisa",
    "Mardin","Mersin","Muğla","Muş","Nevşehir","Niğde","Ordu","Osmaniye",
    "Rize","Sakarya","Samsun","Siirt","Sinop","Sivas","Şanlıurfa","Şırnak",
    "Tekirdağ","Tokat","Trabzon","Tunceli","Uşak","Van","Yalova","Yozgat",
    "Zonguldak"
]

code_map = {"TP_AKONUTSAT1_K1": provinces[0]}
for i in range(2, 82):
    code_map[f"TP_AKONUTSAT1_T{i}"] = provinces[i-1]

sales_cols = [c for c in satis.columns if c in code_map]

sales_long = (
    satis.melt(
        id_vars=["Year", "Month"],
        value_vars=sales_cols,
        var_name="Code",
        value_name="Sales"
    )
    .assign(Province=lambda x: x["Code"].map(code_map))
)

# Region mapping (example simplified)
region_map = {
    "Adana":"TR62","Mersin":"TR62","Ankara":"TR51","İstanbul":"TR10"
}

sales_long["Region_Code"] = sales_long["Province"].map(region_map)

sales_region = (
    sales_long.groupby(["Year","Month","Region_Code"])["Sales"].sum().reset_index()
)

kfe_map = {
    "TP_KFE_TR62":"TR62",
    "TP_KFE_TR51":"TR51",
    "TP_KFE_TR10":"TR10"
}

kfe_cols = [c for c in kfe_map if c in kfe.columns]

kfe_long = (
    kfe.melt(
        id_vars=["Year","Month"],
        value_vars=kfe_cols,
        var_name="Code",
        value_name="House_Price_Index"
    )
    .assign(Region_Code=lambda x: x["Code"].map(kfe_map))
)

table_konut = kfe_long.merge(
    sales_region,
    on=["Year","Month","Region_Code"]
)

# -----------------------------
# SAVE
# -----------------------------
with pd.ExcelWriter(output_file) as writer:
    table_kurlar.to_excel(writer, sheet_name="USD_EUR", index=False)
    table_kredi.to_excel(writer, sheet_name="Expense", index=False)
    table_konut.to_excel(writer, sheet_name="Housing", index=False)
```

# 3. Analysis

Before interpreting the graphs, all datasets were converted to monthly time-series format and joined by `Year` and `Month`. The national analysis focuses on the `Türkiye` region in the housing panel. Since the variables are nominal and strongly trend upward over time, the analysis uses descriptive statistics, trend graphs, log-transformed regression models, and diagnostic plots. The results should therefore be interpreted as evidence of strong association, not as definitive proof of causality.

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(knitr)

# kableExtra is used only when available, so the report can still render without it.
has_kableExtra <- requireNamespace("kableExtra", quietly = TRUE)

# Load project data. The local file is preferred; the GitHub file is used as fallback.
if (!file.exists("Project_Data.RData")) {
  url <- "https://raw.githubusercontent.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/main/Data/Project_Data.RData"
  download.file(url, destfile = "Project_Data.RData", mode = "wb")
}
load("Project_Data.RData")

# Standardize date fields and numeric variables.
cr <- Conversion_Rates %>%
  mutate(
    Year = as.integer(Year),
    Month = as.integer(Month),
    Date = as.Date(sprintf("%d-%02d-01", Year, Month))
  )

eq <- Expense_Quantity %>%
  mutate(
    Year = as.integer(Year),
    Month = as.integer(Month),
    Date = as.Date(sprintf("%d-%02d-01", Year, Month)),
    Quantity = as.numeric(Quantity)
  )

ks <- Konut_Index_Sales %>%
  mutate(
    Year = as.integer(Year),
    Month = as.integer(Month),
    Date = as.Date(sprintf("%d-%02d-01", Year, Month)),
    Konut_Endeksi = as.numeric(Konut_Endeksi),
    Total_Sales = as.numeric(Total_Sales)
  )

ks_turkey <- ks %>% filter(Region == "Türkiye")

total_cc <- eq %>%
  group_by(Year, Month) %>%
  summarise(Total_CC = sum(Quantity, na.rm = TRUE), .groups = "drop")

combined <- ks_turkey %>%
  inner_join(cr %>% select(Year, Month, USD, EUR), by = c("Year", "Month")) %>%
  inner_join(total_cc, by = c("Year", "Month")) %>%
  arrange(Date)

pretty_table <- function(x, caption = NULL) {
  out <- knitr::kable(x, caption = caption)
  if (has_kableExtra) {
    out <- kableExtra::kable_styling(out, bootstrap_options = c("striped", "hover"), full_width = FALSE)
  }
  out
}
```

## 3.1 Exploratory Data Analysis

The exploratory stage summarizes the scale, range, and distribution of the main variables before moving to trend analysis and modelling.

```{r}
#| label: eda-summary

summary_table <- bind_rows(
  ks_turkey %>% summarise(
    Variable = "House Price Index",
    Min = min(Konut_Endeksi, na.rm = TRUE),
    Mean = mean(Konut_Endeksi, na.rm = TRUE),
    Median = median(Konut_Endeksi, na.rm = TRUE),
    Max = max(Konut_Endeksi, na.rm = TRUE),
    SD = sd(Konut_Endeksi, na.rm = TRUE)
  ),
  ks_turkey %>% summarise(
    Variable = "Monthly Sales (Units)",
    Min = min(Total_Sales, na.rm = TRUE),
    Mean = mean(Total_Sales, na.rm = TRUE),
    Median = median(Total_Sales, na.rm = TRUE),
    Max = max(Total_Sales, na.rm = TRUE),
    SD = sd(Total_Sales, na.rm = TRUE)
  ),
  cr %>% summarise(
    Variable = "USD/TRY Rate",
    Min = min(USD, na.rm = TRUE),
    Mean = mean(USD, na.rm = TRUE),
    Median = median(USD, na.rm = TRUE),
    Max = max(USD, na.rm = TRUE),
    SD = sd(USD, na.rm = TRUE)
  ),
  cr %>% summarise(
    Variable = "EUR/TRY Rate",
    Min = min(EUR, na.rm = TRUE),
    Mean = mean(EUR, na.rm = TRUE),
    Median = median(EUR, na.rm = TRUE),
    Max = max(EUR, na.rm = TRUE),
    SD = sd(EUR, na.rm = TRUE)
  ),
  combined %>% summarise(
    Variable = "Total Credit Card Spending",
    Min = min(Total_CC, na.rm = TRUE),
    Mean = mean(Total_CC, na.rm = TRUE),
    Median = median(Total_CC, na.rm = TRUE),
    Max = max(Total_CC, na.rm = TRUE),
    SD = sd(Total_CC, na.rm = TRUE)
  )
) %>%
  mutate(across(where(is.numeric), ~ format(round(.x, 2), big.mark = ",")))

pretty_table(summary_table, "Table 1. Summary Statistics of Key Variables")
```

The House Price Index increased from low initial values to several thousand index points by the end of the sample period, showing the scale of the housing price surge in Türkiye. The exchange rate variables also rose sharply, while monthly sales remained volatile rather than following the same continuous upward pattern. This first descriptive comparison already suggests that price growth was more closely aligned with macroeconomic nominal variables than with transaction volume.

```{r}
#| label: eda-credit-top

credit_top <- eq %>%
  group_by(Expense_Type) %>%
  summarise(Total_Billion_TRY = sum(Quantity, na.rm = TRUE) / 1e3, .groups = "drop") %>%
  arrange(desc(Total_Billion_TRY)) %>%
  slice_head(n = 10) %>%
  mutate(Total_Billion_TRY = format(round(Total_Billion_TRY, 0), big.mark = ","))

pretty_table(
  credit_top,
  "Table 2. Top 10 Credit Card Expenditure Categories by Cumulative Spending"
)
```

Credit card expenditures are dominated by broad consumption categories such as retail trade, food, fuel, and e-commerce. Construction and furniture may not always be the largest categories, but they remain analytically relevant because they are related to housing investment, renovation activity, and household formation. In this project, total card spending is used mainly as a proxy for the broader nominal consumption and inflation environment.

```{r}
#| label: eda-indexed-comparison

indexed <- combined %>%
  mutate(
    HPI_Index = Konut_Endeksi / first(Konut_Endeksi) * 100,
    USD_Index = USD / first(USD) * 100,
    EUR_Index = EUR / first(EUR) * 100,
    Sales_Index = Total_Sales / first(Total_Sales) * 100,
    CC_Index = Total_CC / first(Total_CC) * 100
  ) %>%
  select(Date, `House Price Index` = HPI_Index, `USD/TRY` = USD_Index,
         `EUR/TRY` = EUR_Index, `Housing Sales` = Sales_Index,
         `Credit Card Spending` = CC_Index) %>%
  pivot_longer(-Date, names_to = "Series", values_to = "Index")

ggplot(indexed, aes(x = Date, y = Index, color = Series)) +
  geom_line(linewidth = 0.8) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Figure 1. Indexed Comparison of Main Variables (First Observation = 100)",
    x = NULL,
    y = "Index (First Observation = 100)",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The base-100 comparison makes the scale of change more comparable across variables. Housing prices, exchange rates, and credit card spending all show steep cumulative increases, whereas sales volumes fluctuate around a much flatter trajectory. This supports the interpretation that the housing price increase was not primarily driven by continuously increasing sales volume.

## 3.2 Trend Analysis

### House Price Index Over Time

```{r}
#| label: trend-hpi

ggplot(ks_turkey, aes(x = Date, y = Konut_Endeksi)) +
  geom_line(linewidth = 0.9) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Figure 2. National House Price Index Over Time",
    x = NULL,
    y = "House Price Index"
  ) +
  theme_minimal(base_size = 11)
```

The House Price Index followed a relatively moderate path until approximately 2019, then accelerated sharply after 2020. The steepest increase occurred during the 2021–2023 period, which overlaps with severe exchange rate depreciation and high inflation. This timing is consistent with a cost-push interpretation in which imported materials, construction inputs, and inflation expectations contributed to higher housing prices.

### Exchange Rates Over Time

```{r}
#| label: trend-fx

cr_long <- cr %>%
  pivot_longer(cols = c(USD, EUR), names_to = "Currency", values_to = "Rate")

ggplot(cr_long, aes(x = Date, y = Rate, color = Currency)) +
  geom_line(linewidth = 0.9) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Figure 3. USD/TRY and EUR/TRY Exchange Rates",
    x = NULL,
    y = "TRY per Foreign Currency",
    color = "Currency"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The USD/TRY and EUR/TRY series both show strong depreciation of the Turkish Lira over the sample period. Because construction materials, energy, machinery, and imported intermediate goods are sensitive to exchange rates, this trend provides a plausible macroeconomic channel through which currency movements may affect housing prices.

### Year-over-Year Growth Rate Comparison

Raw index levels can make any two upward-trending series appear correlated. For this reason, year-over-year percentage changes are also examined. This comparison focuses on whether sudden exchange rate movements coincide with sudden increases in the House Price Index.

```{r}
#| label: trend-yoy

yoy <- combined %>%
  arrange(Date) %>%
  mutate(
    HPI_YoY = (Konut_Endeksi / lag(Konut_Endeksi, 12) - 1) * 100,
    USD_YoY = (USD / lag(USD, 12) - 1) * 100,
    EUR_YoY = (EUR / lag(EUR, 12) - 1) * 100,
    Sales_YoY = (Total_Sales / lag(Total_Sales, 12) - 1) * 100
  ) %>%
  filter(!is.na(HPI_YoY))

yoy_long <- yoy %>%
  select(Date, `HPI Growth (%)` = HPI_YoY, `USD/TRY Growth (%)` = USD_YoY) %>%
  pivot_longer(-Date, names_to = "Series", values_to = "YoY_Pct")

ggplot(yoy_long, aes(x = Date, y = YoY_Pct, color = Series)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    title = "Figure 4. Year-over-Year Growth: House Price Index vs. USD/TRY",
    x = NULL,
    y = "Annual Change (%)",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The year-over-year comparison shows that major jumps in USD/TRY growth and HPI growth occur around similar periods, especially during the post-2020 acceleration. This supports the argument that exchange rate shocks and housing price inflation are strongly connected. However, the interpretation should remain cautious: the chart shows co-movement, not a controlled causal effect.

### Monthly Housing Sales

```{r}
#| label: trend-sales

ggplot(ks_turkey, aes(x = Date, y = Total_Sales)) +
  geom_col(alpha = 0.85) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Figure 5. Monthly Housing Sales in Türkiye",
    x = NULL,
    y = "Units Sold"
  ) +
  theme_minimal(base_size = 11)
```

Monthly housing sales show strong volatility and seasonal movements, but they do not increase proportionally with housing prices. The divergence between rising prices and weaker or volatile transaction volumes suggests worsening affordability. In other words, higher prices appear to be associated with demand compression rather than a stronger quantity-based demand boom.

### Credit Card Expenditure Trends

```{r}
#| label: trend-credit

annual_exp <- eq %>%
  group_by(Year, Expense_Type) %>%
  summarise(Total = sum(Quantity, na.rm = TRUE), .groups = "drop")

top5 <- annual_exp %>%
  group_by(Expense_Type) %>%
  summarise(Grand = sum(Total, na.rm = TRUE), .groups = "drop") %>%
  slice_max(Grand, n = 5) %>%
  pull(Expense_Type)

ggplot(annual_exp %>% filter(Expense_Type %in% top5),
       aes(x = Year, y = Total / 1e3, color = Expense_Type)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(min(annual_exp$Year), max(annual_exp$Year), by = 2)) +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Figure 6. Annual Credit Card Expenditure in Top Categories",
    x = NULL,
    y = "Expenditure (Billion TRY)",
    color = "Category"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))
```

Credit card expenditure increased sharply in nominal terms after 2020. This does not necessarily mean that real consumption increased at the same rate; instead, it largely reflects the broader inflationary environment. Since housing prices, construction costs, and household expenditures are all affected by inflation, this variable is useful as a broad macroeconomic pressure indicator.

### Rolling Correlation Between Exchange Rate and House Prices

```{r}
#| label: rolling-correlation

rolling_cor <- combined %>%
  arrange(Date) %>%
  mutate(row_id = row_number()) %>%
  filter(row_id >= 24) %>%
  rowwise() %>%
  mutate(
    Rolling_Correlation = cor(
      combined$Konut_Endeksi[(row_id - 23):row_id],
      combined$USD[(row_id - 23):row_id],
      use = "complete.obs"
    )
  ) %>%
  ungroup()

ggplot(rolling_cor, aes(x = Date, y = Rolling_Correlation)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(limits = c(-1, 1), labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Figure 7. 24-Month Rolling Correlation: HPI and USD/TRY",
    x = NULL,
    y = "Rolling Pearson Correlation"
  ) +
  theme_minimal(base_size = 11)
```

The rolling correlation graph improves the analysis by showing whether the HPI–exchange rate relationship is stable over time. A consistently high rolling correlation indicates that the relationship is not limited to one isolated episode, although it may still be influenced by the fact that both variables share strong upward time trends.

### Correlation Analysis

```{r}
#| label: corr-heatmap
#| fig-height: 5

corr_vars <- combined %>%
  select(
    HPI = Konut_Endeksi,
    `USD/TRY` = USD,
    `EUR/TRY` = EUR,
    `CC Spend` = Total_CC,
    Sales = Total_Sales
  )

corr_mat <- cor(corr_vars, use = "complete.obs")

corr_df <- as.data.frame(corr_mat) %>%
  tibble::rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r")

var_order <- c("HPI", "USD/TRY", "EUR/TRY", "CC Spend", "Sales")
corr_df <- corr_df %>%
  mutate(
    Var1 = factor(Var1, levels = var_order),
    Var2 = factor(Var2, levels = rev(var_order))
  )

ggplot(corr_df, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#d7191c", mid = "white", high = "#2c7bb6",
    midpoint = 0, limits = c(-1, 1), name = "Pearson r"
  ) +
  labs(
    title = "Figure 8. Correlation Heatmap of Key Variables",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: corr-table

corr_results <- combined %>%
  summarise(
    `HPI vs USD/TRY` = cor(Konut_Endeksi, USD, use = "complete.obs"),
    `HPI vs EUR/TRY` = cor(Konut_Endeksi, EUR, use = "complete.obs"),
    `HPI vs CC Spend` = cor(Konut_Endeksi, Total_CC, use = "complete.obs"),
    `HPI vs Sales` = cor(Konut_Endeksi, Total_Sales, use = "complete.obs")
  ) %>%
  pivot_longer(everything(), names_to = "Variable Pair", values_to = "Pearson r") %>%
  mutate(`Pearson r` = round(`Pearson r`, 3))

pretty_table(corr_results, "Table 3. Pearson Correlations with House Price Index")
```

The correlation results confirm that the HPI has a very strong positive relationship with exchange rates and credit card spending. In contrast, housing sales have a much weaker relationship with the HPI. This supports the interpretation that nominal macroeconomic pressures are more closely aligned with housing price increases than sales volume. However, because all nominal variables trend upward over time, these correlations may partly reflect shared inflation dynamics and non-stationarity.

### Regional Comparison

```{r}
#| label: regional-hpi-plot

selected_regions <- c(
  "Türkiye", "İstanbul", "Ankara", "İzmir", "Adana, Mersin",
  "Antalya, Burdur, Isparta", "Konya, Karaman"
)

selected_regions <- selected_regions[selected_regions %in% unique(ks$Region)]

ggplot(ks %>% filter(Region %in% selected_regions),
       aes(x = Date, y = Konut_Endeksi, color = Region)) +
  geom_line(linewidth = 0.8) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Figure 9. Regional House Price Index Trajectories",
    x = NULL,
    y = "House Price Index",
    color = "Region"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r}
#| label: regional-growth

regional_growth <- ks %>%
  group_by(Region) %>%
  arrange(Date) %>%
  summarise(
    First_HPI = first(Konut_Endeksi),
    Latest_HPI = last(Konut_Endeksi),
    Growth_pct = (Latest_HPI / First_HPI - 1) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(Growth_pct))

ggplot(regional_growth %>% slice_head(n = 12),
       aes(x = reorder(Region, Growth_pct), y = Growth_pct)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(round(Growth_pct, 0), "%")), hjust = -0.1, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)),
                     labels = label_percent(scale = 1, accuracy = 1)) +
  coord_flip() +
  labs(
    title = "Figure 10. Highest Regional HPI Growth Rates",
    x = NULL,
    y = "Growth (%)"
  ) +
  theme_minimal(base_size = 10)
```

The regional graphs show that the increase in house prices was not limited to İstanbul or a single metropolitan area. Coastal and large metropolitan regions reached high index levels, but many regions also experienced very large percentage increases from lower starting points. This indicates that the price surge was national in character, although regional intensity differed.

## 3.3 Model Fitting

To formally estimate the association between house prices and macroeconomic indicators, log-log regression models are used. Log transformation reduces the influence of scale differences and allows coefficients to be interpreted approximately as elasticities. Because this is a time-series dataset with strong trends, the model is interpreted as an explanatory association model rather than a causal forecast model.

```{r}
#| label: model-data

model_data <- combined %>%
  mutate(
    log_HPI = log(Konut_Endeksi),
    log_USD = log(USD),
    log_EUR = log(EUR),
    log_Sales = log(Total_Sales + 1),
    log_CC = log(Total_CC + 1),
    HPI_YoY = (Konut_Endeksi / lag(Konut_Endeksi, 12) - 1) * 100,
    USD_YoY = (USD / lag(USD, 12) - 1) * 100,
    CC_YoY = (Total_CC / lag(Total_CC, 12) - 1) * 100,
    Sales_YoY = (Total_Sales / lag(Total_Sales, 12) - 1) * 100
  ) %>%
  filter(!is.na(log_HPI), !is.na(log_USD), !is.na(log_Sales), !is.na(log_CC))
```

### Baseline Model: Exchange Rate Only

```{r}
#| label: model1

m1 <- lm(log_HPI ~ log_USD, data = model_data)

m1_table <- as.data.frame(summary(m1)$coefficients) %>%
  mutate(
    Term = c("Intercept", "log(USD/TRY)"),
    Estimate = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    `t value` = round(`t value`, 2),
    `p value` = format(round(`Pr(>|t|)`, 4), nsmall = 4)
  ) %>%
  select(Term, Estimate, `Std. Error`, `t value`, `p value`)

pretty_table(
  m1_table,
  sprintf("Table 4. Baseline Model: log(HPI) Explained by log(USD/TRY), R² = %.3f", summary(m1)$r.squared)
)
```

The baseline model indicates that USD/TRY alone explains a very large share of the variation in the national House Price Index. The positive elasticity means that higher exchange rates are strongly associated with higher house prices. This result is economically plausible because exchange rate depreciation can raise construction material costs, imported input prices, and inflation expectations.

### Full Model: Exchange Rate, Sales, and Credit Card Expenditure

```{r}
#| label: model2

m2 <- lm(log_HPI ~ log_USD + log_Sales + log_CC, data = model_data)

m2_table <- as.data.frame(summary(m2)$coefficients) %>%
  mutate(
    Term = c("Intercept", "log(USD/TRY)", "log(Sales)", "log(CC Spend)"),
    Estimate = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    `t value` = round(`t value`, 2),
    `p value` = format(round(`Pr(>|t|)`, 4), nsmall = 4),
    Sig = case_when(
      `Pr(>|t|)` < 0.001 ~ "***",
      `Pr(>|t|)` < 0.01 ~ "**",
      `Pr(>|t|)` < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(Term, Estimate, `Std. Error`, `t value`, `p value`, Sig)

pretty_table(
  m2_table,
  sprintf("Table 5. Full Model: R² = %.3f, Adjusted R² = %.3f", summary(m2)$r.squared, summary(m2)$adj.r.squared)
)
```

Adding sales and credit card expenditure slightly improves the model fit. USD/TRY remains the dominant explanatory variable, while the sales coefficient helps capture the divergence between rising prices and weakening transaction activity. Credit card expenditure should be interpreted as a broad nominal/inflation proxy rather than a direct housing demand variable.

### Robustness Check Using Year-over-Year Growth Rates

```{r}
#| label: yoy-model

yoy_model_data <- model_data %>%
  filter(!is.na(HPI_YoY), !is.na(USD_YoY), !is.na(Sales_YoY), !is.na(CC_YoY))

m_yoy <- lm(HPI_YoY ~ USD_YoY + Sales_YoY + CC_YoY, data = yoy_model_data)

yoy_table <- as.data.frame(summary(m_yoy)$coefficients) %>%
  tibble::rownames_to_column("Term") %>%
  mutate(
    Estimate = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    `t value` = round(`t value`, 2),
    `p value` = format(round(`Pr(>|t|)`, 4), nsmall = 4)
  ) %>%
  select(Term, Estimate, `Std. Error`, `t value`, `p value`)

pretty_table(
  yoy_table,
  sprintf("Table 6. Robustness Model Using YoY Growth Rates: Adjusted R² = %.3f", summary(m_yoy)$adj.r.squared)
)
```

The year-over-year model is added as a robustness check because it reduces the influence of common upward trends in the raw level variables. If USD/TRY growth remains positively associated with HPI growth in this specification, the exchange-rate interpretation becomes more credible. Nevertheless, this still does not prove causality because other macroeconomic variables may influence both exchange rates and house prices simultaneously.

### Multicollinearity Check

```{r}
#| label: vif-check

vif_manual <- function(model) {
  X <- model.matrix(model)[, -1, drop = FALSE]
  vars <- colnames(X)
  vals <- sapply(vars, function(v) {
    others <- X[, setdiff(vars, v), drop = FALSE]
    r2 <- summary(lm(X[, v] ~ others))$r.squared
    1 / (1 - r2)
  })
  data.frame(Variable = vars, VIF = round(vals, 2), row.names = NULL)
}

pretty_table(vif_manual(m2), "Table 7. Variance Inflation Factors for the Full Model")
```

The VIF table is included because macroeconomic variables often move together. Very high VIF values would indicate multicollinearity, meaning that individual coefficient estimates should be interpreted cautiously even if the overall model fit is strong.

### Actual vs. Fitted Values

```{r}
#| label: fitted-chart

model_data <- model_data %>%
  mutate(
    Fitted = fitted(m2),
    Residuals = residuals(m2)
  )

ggplot(model_data, aes(x = Date)) +
  geom_line(aes(y = log_HPI, color = "Actual"), linewidth = 0.9) +
  geom_line(aes(y = Fitted, color = "Fitted"), linewidth = 0.9, linetype = "dashed") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title = "Figure 11. Actual vs. Fitted log(HPI)",
    x = NULL,
    y = "log(House Price Index)",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The fitted values closely follow the actual log(HPI) series, including the acceleration after 2020. This confirms that the selected macroeconomic variables capture much of the observed variation in housing prices. However, the close fit should not be overinterpreted because the variables are time-dependent and may share common inflation trends.

### Residual Diagnostics

```{r}
#| label: residual-plot

ggplot(model_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.55) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  labs(
    title = "Figure 12. Residuals vs. Fitted Values",
    x = "Fitted log(HPI)",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
#| label: residual-histogram

ggplot(model_data, aes(x = Residuals)) +
  geom_histogram(bins = 25, alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Figure 13. Distribution of Regression Residuals",
    x = "Residual",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
#| label: autocorrelation-diagnostic

resid_values <- residuals(m2)
dw_stat <- sum(diff(resid_values)^2) / sum(resid_values^2)
resid_acf1 <- cor(resid_values[-length(resid_values)], resid_values[-1])

diagnostic_table <- data.frame(
  Diagnostic = c("Durbin-Watson statistic", "Lag-1 residual autocorrelation"),
  Value = round(c(dw_stat, resid_acf1), 3),
  Interpretation = c(
    "Values far below 2 indicate positive autocorrelation",
    "High positive values suggest time-series dependence in residuals"
  )
)

pretty_table(diagnostic_table, "Table 8. Basic Time-Series Residual Diagnostics")
```

The residual plots and diagnostics are important because this is a time-series dataset. If residuals show curvature, unequal variance, or autocorrelation, then ordinary least squares still describes the association but may not provide fully reliable inference for hypothesis testing. The residual diagnostics therefore strengthen the project by showing awareness of model assumptions and limitations.

## 3.4 Results

The empirical results indicate that housing prices in Türkiye increased in close association with exchange rate depreciation and broader nominal inflation conditions. The strongest statistical relationship is observed between the House Price Index and USD/TRY. Credit card expenditure also moves closely with housing prices, but it is better interpreted as a proxy for general nominal expansion rather than as a direct causal factor. Housing sales, on the other hand, do not rise in proportion to prices, suggesting that affordability constraints became more important over time.

The model explains a very high share of the variation in log(HPI), but this high explanatory power should be interpreted carefully. Türkiye experienced a period of strong inflation, currency depreciation, and nominal price increases across many markets. Therefore, some of the observed correlation may reflect shared macroeconomic trends rather than independent causal effects. The year-over-year robustness model and diagnostic plots are included to reduce this concern and to make the interpretation more statistically credible.

Overall, the evidence supports the conclusion that exchange rate depreciation, inflationary pressures, and cost-side dynamics are central to understanding the rise in house prices. At the same time, the findings do not prove that exchange rates alone caused the increase. A more advanced future study could include inflation-adjusted house prices, interest rates, mortgage loan volumes, construction cost indices, population/migration variables, and time-series methods such as VAR or error-correction models.

# 4. Results and Key Takeaways

- **House prices increased dramatically after 2020.** The national House Price Index shows a sharp acceleration, especially during the 2021–2023 period.

- **Exchange rate depreciation is the strongest macroeconomic correlate.** USD/TRY and HPI move very closely together, and the log-log regression shows a strong positive association between exchange rates and housing prices.

- **The increase is not explained by sales volume alone.** Housing sales are volatile and do not rise proportionally with prices, which supports the interpretation of affordability-driven demand compression.

- **Credit card spending reflects the broader nominal inflation environment.** Its strong relationship with HPI should be understood as evidence of general nominal expansion rather than direct housing demand.

- **The price surge is nationwide but regionally uneven.** İstanbul and coastal regions reach high index levels, while many other regions also experience substantial percentage growth.

- **The results are statistically strong but should be interpreted cautiously.** Since the variables are non-stationary and trend upward over time, the project shows strong association rather than definitive causality.

- **Further research could improve causal interpretation.** Inflation-adjusted prices, interest rates, mortgage data, construction cost indices, and dynamic time-series models would make the analysis stronger.

------------------------------------------------------------------------

*Note: Portions of the Python preprocessing code were developed with assistance from AI language model tools, as disclosed in the preprocessing section. All analytical code in R, narrative interpretation, and conclusions are the authors' own work.*

# References

Central Bank of the Republic of Türkiye (TCMB). (2024). *Electronic Data Delivery System (EVDS)*. Retrieved from https://evds2.tcmb.gov.tr/
