---
title: "Why Have House Prices in Turkey Gone Up in the Last Ten Years?"
author:
  - name: "Ömer Furkan Demirci"
  - name: "Fatih Alper Vural"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    fig-width: 9
    fig-height: 5
    fig-cap-location: bottom
    df-print: paged
  pdf:
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    margin-left: 2cm
    margin-right: 2cm
    margin-top: 2cm
    margin-bottom: 2cm
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Abstract {.unnumbered}

Housing affordability has emerged as one of the most pressing socioeconomic challenges in Türkiye over the past decade. This study investigates the key drivers of the sharp and sustained rise in housing prices between 2014 and 2024, drawing on four official monthly time-series datasets obtained from the Central Bank of the Republic of Türkiye (TCMB/EVDS): the House Price Index (HPI), housing sales statistics, USD/EUR exchange rates, and credit card sectoral expenditure data. Through exploratory analysis, correlation testing, regional comparison, and multiple linear regression modelling, we find that exchange rate depreciation and broad consumer spending inflation are the dominant macroeconomic forces behind rising house prices, while sales volumes display an inverse relationship with price levels — consistent with demand compression at high price levels. Regional heterogeneity is substantial: İstanbul and western metropolitan areas lead price growth, whereas interior Anatolian regions lag considerably. Our model explains approximately 95% of the variance in the national HPI, suggesting that macroeconomic indicators are strong predictors of housing price trajectories in Türkiye.

------------------------------------------------------------------------

# Project Overview and Scope

We want to look into the following question for this project:\
\
**Why have house prices in Turkey gone up in the last ten years?**\
\
To answer this question, we made a dataset using four monthly time-series datasets from the Central Bank of the Republic of Türkiye's (TCMB) Electronic Data Delivery System (EVDS). These datasets have:\

-   Housing Price Index

-   Sales of Houses

-   Exchange Rates (USD/TRY and EUR/TRY)

-   Credit Card Sectoral Expenditures\
    \

We chose this topic because the cost of housing has become one of the most important social and economic issues in Türkiye. It is important to know what caused this rise in order to make good decisions about policy and the economy. This problem is also a good fit for a data analytics approach because it needs to combine multiple datasets, do preprocessing, and look at how things change over time and in different areas.

We want to find the main things that cause housing prices to go up. Specifically, we want to look at:

-   How exchange rates affect the price of homes

-   The link between changes in housing prices and demand (sales)

-   The function of consumption behavior (credit card expenditure) as an indicator of demand conditions.

-   Differences in how housing works in different regions

# Data

You can see Our data in R Data format below.

[Project Data](https://github.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/blob/e273d93492dc0a2ea56f85a472a398fe4909c2d7/Data/Project_Data.RData)

```{r}
url <- "https://raw.githubusercontent.com/emu-hacettepe-analytics/emu660-spring2026-ofdemirci/main/Data/Project_Data.RData"
download.file(url, destfile = "Project_Data.RData", mode = "wb")
load("Project_Data.RData")

ls()
```

## Data Source

The data used in this project were obtained from the **Electronic Data Delivery System (EVDS)** of the Central Bank of the Republic of Türkiye (TCMB). EVDS provides official, high-frequency economic data and is widely used for academic and professional analysis.

The project is based on four datasets:

-   **House Price Index (HPI)**\
    Monthly housing price index at national and regional levels

-   **House Sales Statistics**\
    Monthly housing sales data at the provincial level

-   **Exchange Rates (USD, EUR)**\
    Monthly exchange rate data

-   **Credit Card Sectoral Expenditures**\
    Monthly sector-based credit card spending data

## General Information About Data

The processed dataset consists of three main analytical tables:

### Exchange Rate Table

-   **Description**:\
    This table contains the columns Year, Month, USD Conversion Rate, and EUR Conversion Rate. Represents monthly exchange rate movements. These variables are used as indicators of macroeconomic conditions and cost-side pressures that may affect housing prices.

-   In general, the exchange rate variables show a strong upward trend over time rather than a stable distribution.

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(scales)
    usd_eur <- Conversion_Rates %>%
      filter(Year < 2026) %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(usd_eur, aes(x = Date)) +
      geom_line(aes(y = USD, color = "USD")) +
      geom_line(aes(y = EUR, color = "EUR")) +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      scale_y_continuous(labels = label_number(accuracy = 0.1)) +
      labs(
        title = "Exchange Rates Over Time",
        x = NULL,
        y = "TRY per Foreign Currency",
        color = "Currency"
      ) +
      theme_minimal()
    ```

### Credit Card Expenditure Table

-   **Description**:

-   This table contains Year, Month, Expense Type, and Quantity.\

-   Expense Type identifies the spending category, and Quantity gives the corresponding spending amount for that month.

-   This table contains sectoral spending amounts derived from credit card transactions.

-   The values in this table are expected to differ substantially across categories. Some categories such as retail, food, fuel, and e-commerce are likely to have much larger values than others.

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(scales)
    expense_quantity <- Expense_Quantity %>%
      filter(Year < 2026) %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(
      expense_quantity %>%
        filter(Expense_Type %in% c("E-Alışveriş", "Çeşitli Gıda", "Giyim ve Aksesuar", "Benzin ve Yakıt İstasyonları")),
      aes(x = Date, y = Quantity / 1e3, color = Expense_Type)
    ) +
      geom_line() +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      scale_y_continuous(labels = label_comma(accuracy = 1)) +
      labs(
        title = "Credit Card Expenditures for Selected Categories",
        x = NULL,
        y = "Spending (Billion TRY)",
        color = "Expense Type"
      ) +
      theme_minimal() +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8))
    ```

### Housing Panel Dataset

-   **Description**:\
    This is the main table for the project and contains Year, Month, Region, House Price Index, and Total Sales.

    Region represents the housing price index regions as in NUTS2 level, House Price Index shows the monthly house price index for that region, and Total Sales gives the aggregated monthly house sales value for the same region.

    ```{r}
    library(dplyr)
    library(ggplot2)
    library(scales)

    konut_index_sales <- Konut_Index_Sales %>%
      filter(Year < 2026) %>%
      mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

    ggplot(
      konut_index_sales %>% filter(Region == "Türkiye"),
      aes(x = Date, y = Konut_Endeksi)
    ) +
      geom_line() +
      scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
      scale_y_continuous(labels = label_comma(accuracy = 1)) +
      labs(
        title = "House Price Index in Türkiye Over Time",
        x = NULL,
        y = "House Price Index"
      ) +
      theme_minimal()
    ```

```{r}
library(dplyr)
library(ggplot2)
library(scales)

ggplot(
  Konut_Index_Sales %>%
    filter(Year < 2026) %>%
    mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month))) %>%
    filter(Region %in% c("Türkiye", "İstanbul", "İzmir", "Ankara", "Adana, Mersin")),
  aes(x = Date, y = Total_Sales, color = Region)
) +
  geom_line() +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(
    title = "Total House Sales for Selected Regions",
    x = NULL,
    y = "Units Sold",
    color = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Potential Analyses

-   time-series plots of house prices, exchange rates, and sales,
-   regional comparisons of house price growth,
-   correlations between exchange rates and the house price index,
-   relationships between sales volume and housing prices,
-   category-based spending trends over time,

## Reason of Choice

Housing prices are one of the most important economic and social issues in Türkiye. The rapid increase in housing prices has significant implications for affordability, investment decisions, and economic stability.

All datasets are:

-   Consistent in frequency (monthly)

-   Long enough for trend analysis

-   Rich enough for regional comparison

    This makes it easy for analyzing.

## Preprocessing

The raw EVDS datasets were provided in Excel format and contained different structures (wide vs. long) and aggregation levels (province vs. region). We performed preprocessing in Python using pandas to standardize date formats, reshape wide datasets into long format, map categorical codes to meaningful labels, and align all datasets at a consistent monthly and regional level. In particular, house sales data were converted from province-level columns to regional aggregates compatible with the House Price Index. The final result consists of three clean analytical tables ready for further analysis.

Another key issue was that EVDS datasets use **coded column names** (e.g., `KT2`, `T58`, `TR62`) instead of descriptive labels. To make the data interpretable and consistent across sources, we created mapping dictionaries in Python to translate these codes into meaningful categories, provinces, and regions.

Here you can find our python code.\
\

``` py
# We explained the data and the output tables to AI. It gave us this code's first version. Then we fixed it according to our needs
import pandas as pd
from pathlib import Path

downloads = Path.home() / "Downloads"

# Files
kurlar_file = downloads / "Kurlar.xlsx"
kredi_file = downloads / "Kredi Harcama.xlsx"
konut_satis_file = downloads / "Konut Satış.xlsx"
konut_fiyat_file = downloads / "Konut Fiyat Endexi.xlsx"

output_file = downloads / "evds_preprocessed_tables_fixed.xlsx"

# -----------------------------
# Helper
# -----------------------------
def split_year_month(df):
    df["Tarih"] = pd.to_datetime(df["Tarih"], format="%Y-%m")
    df["Year"] = df["Tarih"].dt.year
    df["Month"] = df["Tarih"].dt.month
    return df

# -----------------------------
# Read
# -----------------------------
kurlar = split_year_month(pd.read_excel(kurlar_file))
kredi = split_year_month(pd.read_excel(kredi_file))
satis = split_year_month(pd.read_excel(konut_satis_file))
kfe = split_year_month(pd.read_excel(konut_fiyat_file))

# -----------------------------
# TABLE 1: Exchange Rates
# -----------------------------
table_kurlar = (
    kurlar[["Year", "Month", "TP_DK_USD_S_YTL", "TP_DK_EUR_S_YTL"]]
    .rename(columns={
        "TP_DK_USD_S_YTL": "USD",
        "TP_DK_EUR_S_YTL": "EUR"
    })
)

# -----------------------------
# TABLE 2: Credit Expenditures
# -----------------------------
expense_type_map = {
    "TP_KKHARTUT_KT2": "Car Rental",
    "TP_KKHARTUT_KT3": "Car Sales/Service",
    "TP_KKHARTUT_KT4": "Fuel",
    "TP_KKHARTUT_KT5": "Food",
    "TP_KKHARTUT_KT6": "Direct Marketing",
    "TP_KKHARTUT_KT7": "Education",
    "TP_KKHARTUT_KT8": "Electronics",
    "TP_KKHARTUT_KT9": "Clothing",
    "TP_KKHARTUT_KT10": "Airlines",
    "TP_KKHARTUT_KT11": "Services",
    "TP_KKHARTUT_KT12": "Accommodation",
    "TP_KKHARTUT_KT13": "Social",
    "TP_KKHARTUT_KT14": "Entertainment",
    "TP_KKHARTUT_KT15": "Jewelry",
    "TP_KKHARTUT_KT16": "Retail",
    "TP_KKHARTUT_KT17": "Furniture",
    "TP_KKHARTUT_KT18": "Construction",
    "TP_KKHARTUT_KT19": "Health",
    "TP_KKHARTUT_KT20": "Travel",
    "TP_KKHARTUT_KT21": "Insurance",
    "TP_KKHARTUT_KT22": "Telecom",
    "TP_KKHARTUT_KT23": "Hardware",
    "TP_KKHARTUT_KT24": "Food Service",
    "TP_KKHARTUT_KT25": "Misc",
    "TP_KKHARTUT_KT26": "Other",
    "TP_KKHARTUT_KT50": "E-Commerce"
}

kredi_cols = [c for c in expense_type_map if c in kredi.columns]

table_kredi = (
    kredi.melt(
        id_vars=["Year", "Month"],
        value_vars=kredi_cols,
        var_name="Code",
        value_name="Quantity"
    )
    .assign(Expense_Type=lambda x: x["Code"].map(expense_type_map))
    [["Year", "Month", "Expense_Type", "Quantity"]]
)

# -----------------------------
# TABLE 3: Housing Panel
# -----------------------------
provinces = [
    "Adana","Adıyaman","Afyonkarahisar","Ağrı","Aksaray","Amasya","Ankara",
    "Antalya","Ardahan","Artvin","Aydın","Balıkesir","Bartın","Batman",
    "Bayburt","Bilecik","Bingöl","Bitlis","Bolu","Burdur","Bursa","Çanakkale",
    "Çankırı","Çorum","Denizli","Diyarbakır","Düzce","Edirne","Elazığ",
    "Erzincan","Erzurum","Eskişehir","Gaziantep","Giresun","Gümüşhane",
    "Hakkâri","Hatay","Iğdır","Isparta","İstanbul","İzmir","Kahramanmaraş",
    "Karabük","Karaman","Kars","Kastamonu","Kayseri","Kırıkkale","Kırklareli",
    "Kırşehir","Kilis","Kocaeli","Konya","Kütahya","Malatya","Manisa",
    "Mardin","Mersin","Muğla","Muş","Nevşehir","Niğde","Ordu","Osmaniye",
    "Rize","Sakarya","Samsun","Siirt","Sinop","Sivas","Şanlıurfa","Şırnak",
    "Tekirdağ","Tokat","Trabzon","Tunceli","Uşak","Van","Yalova","Yozgat",
    "Zonguldak"
]

code_map = {"TP_AKONUTSAT1_K1": provinces[0]}
for i in range(2, 82):
    code_map[f"TP_AKONUTSAT1_T{i}"] = provinces[i-1]

sales_cols = [c for c in satis.columns if c in code_map]

sales_long = (
    satis.melt(
        id_vars=["Year", "Month"],
        value_vars=sales_cols,
        var_name="Code",
        value_name="Sales"
    )
    .assign(Province=lambda x: x["Code"].map(code_map))
)

# Region mapping (example simplified)
region_map = {
    "Adana":"TR62","Mersin":"TR62","Ankara":"TR51","İstanbul":"TR10"
}

sales_long["Region_Code"] = sales_long["Province"].map(region_map)

sales_region = (
    sales_long.groupby(["Year","Month","Region_Code"])["Sales"].sum().reset_index()
)

kfe_map = {
    "TP_KFE_TR62":"TR62",
    "TP_KFE_TR51":"TR51",
    "TP_KFE_TR10":"TR10"
}

kfe_cols = [c for c in kfe_map if c in kfe.columns]

kfe_long = (
    kfe.melt(
        id_vars=["Year","Month"],
        value_vars=kfe_cols,
        var_name="Code",
        value_name="House_Price_Index"
    )
    .assign(Region_Code=lambda x: x["Code"].map(kfe_map))
)

table_konut = kfe_long.merge(
    sales_region,
    on=["Year","Month","Region_Code"]
)

# -----------------------------
# SAVE
# -----------------------------
with pd.ExcelWriter(output_file) as writer:
    table_kurlar.to_excel(writer, sheet_name="USD_EUR", index=False)
    table_kredi.to_excel(writer, sheet_name="Expense", index=False)
    table_konut.to_excel(writer, sheet_name="Housing", index=False)
```

# Analysis

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(knitr)
library(kableExtra)

load("Project_Data.RData")

cr <- Conversion_Rates %>%
  filter(Year < 2026) %>%
  mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

eq <- Expense_Quantity %>%
  filter(Year < 2026) %>%
  mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

ks <- Konut_Index_Sales %>%
  filter(Year < 2026) %>%
  mutate(Date = as.Date(sprintf("%d-%02d-01", Year, Month)))

ks_turkey <- ks %>% filter(Region == "Türkiye")

# Build the shared joined dataset used throughout Analysis
total_cc <- eq %>%
  group_by(Year, Month) %>%
  summarise(Total_CC = sum(Quantity, na.rm = TRUE), .groups = "drop")

combined <- ks_turkey %>%
  inner_join(cr %>% select(Year, Month, USD, EUR), by = c("Year", "Month")) %>%
  inner_join(total_cc, by = c("Year", "Month"))
```

## Exploratory Data Analysis

We begin by examining the basic descriptive statistics of each key variable to understand their range and scale before looking at relationships.

```{r}
#| label: eda-summary

bind_rows(
  ks_turkey %>% summarise(
    Variable = "House Price Index",
    Min      = min(Konut_Endeksi, na.rm = TRUE),
    Mean     = mean(Konut_Endeksi, na.rm = TRUE),
    Median   = median(Konut_Endeksi, na.rm = TRUE),
    Max      = max(Konut_Endeksi, na.rm = TRUE),
    SD       = sd(Konut_Endeksi, na.rm = TRUE)
  ),
  ks_turkey %>% summarise(
    Variable = "Monthly Sales (units)",
    Min      = min(Total_Sales, na.rm = TRUE),
    Mean     = mean(Total_Sales, na.rm = TRUE),
    Median   = median(Total_Sales, na.rm = TRUE),
    Max      = max(Total_Sales, na.rm = TRUE),
    SD       = sd(Total_Sales, na.rm = TRUE)
  ),
  cr %>% summarise(
    Variable = "USD/TRY Rate",
    Min      = min(USD, na.rm = TRUE),
    Mean     = mean(USD, na.rm = TRUE),
    Median   = median(USD, na.rm = TRUE),
    Max      = max(USD, na.rm = TRUE),
    SD       = sd(USD, na.rm = TRUE)
  ),
  cr %>% summarise(
    Variable = "EUR/TRY Rate",
    Min      = min(EUR, na.rm = TRUE),
    Mean     = mean(EUR, na.rm = TRUE),
    Median   = median(EUR, na.rm = TRUE),
    Max      = max(EUR, na.rm = TRUE),
    SD       = sd(EUR, na.rm = TRUE)
  )
) %>%
  mutate(across(where(is.numeric), ~ format(round(.x, 1), big.mark = ","))) %>%
  kable(caption = "Table 1. Summary Statistics of Key Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The HPI has grown from around 100 to several thousand, reflecting the steep price escalation over the decade. The USD/TRY rate rose from roughly 2 to over 30, and monthly sales volumes are in the tens of thousands but show high variability.

```{r}
#| label: eda-credit-top

eq %>%
  group_by(Expense_Type) %>%
  summarise(Total_Billion_TRY = sum(Quantity, na.rm = TRUE) / 1e3, .groups = "drop") %>%
  arrange(desc(Total_Billion_TRY)) %>%
  slice_head(n = 8) %>%
  mutate(Total_Billion_TRY = format(round(Total_Billion_TRY, 0), big.mark = ",")) %>%
  kable(col.names = c("Expense Category", "Total Spending (Billion TRY)"),
        caption = "Table 2. Top 8 Credit Card Expenditure Categories – Cumulative Total") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

Retail trade, food, and e-commerce dominate credit card spending. Construction and furniture categories, while smaller, are directly linked to housing demand.

## Trend Analysis

### House Price Index Over Time

```{r}
#| label: trend-hpi

ggplot(ks_turkey, aes(x = Date, y = Konut_Endeksi)) +
  geom_line(color = "#2c7bb6", linewidth = 0.9) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(title = "Figure 1. National House Price Index Over Time",
       x = NULL, y = "House Price Index") +
  theme_minimal(base_size = 11)
```

The HPI grew modestly through 2019, then accelerated sharply from 2020 onward, with the steepest rise concentrated between 2021 and 2023 — precisely mirroring the most severe TRY depreciation episodes.

### Year-over-Year Growth Rate Comparison

Raw index levels can make any two upward-trending series appear correlated. Year-over-year (YoY) percentage changes are a more meaningful comparison: they show whether **sudden currency shocks coincide with sudden house-price jumps**, which is a much stronger analytical claim.

```{r}
#| label: trend-yoy

yoy <- combined %>%
  arrange(Year, Month) %>%
  mutate(
    HPI_YoY = (Konut_Endeksi / lag(Konut_Endeksi, 12) - 1) * 100,
    USD_YoY = (USD / lag(USD, 12) - 1) * 100
  ) %>%
  filter(!is.na(HPI_YoY))

yoy_long <- yoy %>%
  select(Date, `HPI Growth (%)` = HPI_YoY, `USD/TRY Growth (%)` = USD_YoY) %>%
  pivot_longer(-Date, names_to = "Series", values_to = "YoY_Pct")

ggplot(yoy_long, aes(x = Date, y = YoY_Pct, color = Series)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_color_manual(values = c("HPI Growth (%)" = "#2c7bb6",
                                "USD/TRY Growth (%)" = "#d7191c")) +
  labs(title = "Figure 2. Year-over-Year Growth: House Price Index vs. USD/TRY",
       x = NULL, y = "Annual Change (%)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The YoY chart is more telling than raw levels: the two series spike together, particularly in 2018, 2021, and 2022. When the exchange rate surged by over 80% in 2021–2022, house price growth followed almost immediately — consistent with a direct cost pass-through mechanism.

### Exchange Rates Over Time

```{r}
#| label: trend-fx

cr_long <- cr %>%
  pivot_longer(cols = c(USD, EUR), names_to = "Currency", values_to = "Rate")

ggplot(cr_long, aes(x = Date, y = Rate, color = Currency)) +
  geom_line(linewidth = 0.9) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  scale_color_manual(values = c("USD" = "#d7191c", "EUR" = "#1a9641")) +
  labs(title = "Figure 3. USD/TRY and EUR/TRY Exchange Rates",
       x = NULL, y = "TRY per Foreign Currency", color = "Currency") +
  theme_minimal(base_size = 11)
```

### Annual Housing Sales

```{r}
#| label: trend-sales

annual_sales <- ks_turkey %>%
  group_by(Year) %>%
  summarise(Annual_Sales = sum(Total_Sales, na.rm = TRUE), .groups = "drop")

ggplot(annual_sales, aes(x = Year, y = Annual_Sales)) +
  geom_line(color = "#74add1", linewidth = 1) +
  geom_point(color = "#2c7bb6", size = 3) +
  geom_text(aes(label = format(round(Annual_Sales / 1e3), big.mark = ",")),
            vjust = -1, size = 3, color = "grey30") +
  scale_x_continuous(breaks = seq(2014, 2025, by = 1)) +
  scale_y_continuous(labels = label_comma(accuracy = 1),
                     expand = expansion(mult = c(0.05, 0.15))) +
  labs(title = "Figure 4. Annual Housing Sales – Türkiye",
       subtitle = "Labels show thousands of units sold",
       x = NULL, y = "Units Sold") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Annual sales peaked around 2021 and have trended downward since, even as prices continued to rise sharply — a clear sign of affordability-driven demand compression rather than a demand-led price surge.

### Annual Credit Card Expenditure

```{r}
#| label: trend-credit

annual_exp <- eq %>%
  group_by(Year, Expense_Type) %>%
  summarise(Total = sum(Quantity, na.rm = TRUE), .groups = "drop")

top5 <- annual_exp %>%
  group_by(Expense_Type) %>%
  summarise(Grand = sum(Total), .groups = "drop") %>%
  slice_max(Grand, n = 5) %>%
  pull(Expense_Type)

ggplot(annual_exp %>% filter(Expense_Type %in% top5),
       aes(x = Year, y = Total / 1e3, color = Expense_Type)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2014, 2024, by = 2)) +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(title = "Figure 5. Annual Credit Card Expenditure – Top 5 Categories",
       x = NULL, y = "Expenditure (Billion TRY)", color = "Category") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

All categories grew sharply in nominal terms from 2020 onward, reflecting the broad inflationary environment which also raises construction costs, labour, and land values.

## Correlation Analysis

### Correlation Heatmap

A correlation heatmap summarises the pairwise relationships between all key variables at once, providing a compact picture that makes the regression results easier to interpret.

```{r}
#| label: corr-heatmap
#| fig-height: 5

corr_vars <- combined %>%
  select(
    HPI        = Konut_Endeksi,
    `USD/TRY`  = USD,
    `EUR/TRY`  = EUR,
    `CC Spend` = Total_CC,
    Sales      = Total_Sales
  )

corr_mat <- cor(corr_vars, use = "complete.obs")

# Pivot to long — keep SAME order for both axes so diagonal is top-left to bottom-right
var_order <- c("HPI", "USD/TRY", "EUR/TRY", "CC Spend", "Sales")

corr_df <- as.data.frame(corr_mat) %>%
  tibble::rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r") %>%
  mutate(
    Var1 = factor(Var1, levels = var_order),
    Var2 = factor(Var2, levels = var_order)
  )

ggplot(corr_df, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.5, fontface = "bold") +
  scale_fill_gradient2(low = "#d7191c", mid = "white", high = "#2c7bb6",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Pearson r") +
  scale_y_discrete(limits = rev(var_order)) +
  labs(title = "Figure 6. Correlation Heatmap of Key Variables",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: corr-table

corr_results <- combined %>%
  summarise(
    `HPI vs USD/TRY`  = cor(Konut_Endeksi, USD,          use = "complete.obs"),
    `HPI vs EUR/TRY`  = cor(Konut_Endeksi, EUR,          use = "complete.obs"),
    `HPI vs CC Spend` = cor(Konut_Endeksi, Total_CC,     use = "complete.obs"),
    `HPI vs Sales`    = cor(Konut_Endeksi, Total_Sales,   use = "complete.obs")
  ) %>%
  pivot_longer(everything(), names_to = "Variable Pair", values_to = "Pearson r") %>%
  mutate(`Pearson r` = round(`Pearson r`, 3))

kable(corr_results,
      caption = "Table 3. Pearson Correlations with House Price Index") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The heatmap confirms the narrative: HPI correlates near-perfectly with both exchange rates (r \> 0.97) and credit card expenditure (r \> 0.95), all driven by the same underlying monetary inflation. The HPI–Sales correlation is weakly negative, reflecting demand compression at high price levels.

### Scatter Plot: HPI vs. USD/TRY

Time-series charts can visually exaggerate relationships because both series simply trend upward. A scatter plot directly tests whether the statistical relationship holds across individual observations.

```{r}
#| label: scatter-hpi-usd

ggplot(combined, aes(x = log(USD), y = log(Konut_Endeksi))) +
  geom_point(alpha = 0.5, color = "#2c7bb6") +
  geom_smooth(method = "lm", color = "#d7191c", se = TRUE) +
  scale_x_continuous(labels = label_number(accuracy = 0.1)) +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(title = "Figure 7. log(HPI) vs. log(USD/TRY) – Each Point is One Month",
       x = "log(USD/TRY Rate)", y = "log(House Price Index)") +
  theme_minimal(base_size = 11)
```

The scatter plot on log-transformed variables shows an extremely tight linear relationship with very little scatter, confirming that the correlation is not just a shared time trend artefact — the relationship is genuinely strong at all levels of the exchange rate.

## Regional Comparison

### Regional HPI Trajectories

```{r}
#| label: regional-hpi-plot

# Dynamically find the top-growth region (excluding Türkiye national aggregate)
top_growth_region <- ks %>%
  filter(Region != "Türkiye") %>%
  group_by(Region) %>%
  arrange(Date) %>%
  summarise(
    Growth_pct = (last(Konut_Endeksi) / first(Konut_Endeksi) - 1) * 100,
    .groups = "drop"
  ) %>%
  slice_max(Growth_pct, n = 1) %>%
  pull(Region)

selected_regions <- unique(c(
  "Türkiye", "İstanbul", "Ankara",
  "Adana, Mersin", "Antalya, Burdur, Isparta",
  "Konya, Karaman",
  top_growth_region
))

ggplot(ks %>% filter(Region %in% selected_regions),
       aes(x = Date, y = Konut_Endeksi, color = Region)) +
  geom_line(linewidth = 0.8) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = label_comma(accuracy = 1)) +
  labs(title = "Figure 8. House Price Index by Region",
       x = NULL, y = "House Price Index", color = "Region") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

### Regional Growth Ranking

A ranked bar chart makes the growth comparison across all regions immediately interpretable — far clearer than overlapping line charts.

```{r}
#| label: regional-bar
#| fig-height: 7

region_growth <- ks %>%
  filter(Region != "Türkiye") %>%
  group_by(Region) %>%
  arrange(Date) %>%
  summarise(
    Growth_pct = (last(Konut_Endeksi) / first(Konut_Endeksi) - 1) * 100,
    .groups = "drop"
  )

ggplot(region_growth,
       aes(x = reorder(Region, Growth_pct), y = Growth_pct, fill = Growth_pct)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Growth_pct, 0), "%")),
            hjust = -0.1, size = 3) +
  scale_fill_gradient(low = "#abd9e9", high = "#08306b") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)),
                     labels = label_percent(scale = 1, accuracy = 1)) +
  coord_flip() +
  labs(title = "Figure 9. Total HPI Growth by Region (First to Latest Observation)",
       x = NULL, y = "Growth (%)") +
  theme_minimal(base_size = 10)
```

```{r}
#| label: regional-growth-table

ks %>%
  group_by(Region) %>%
  arrange(Date) %>%
  summarise(
    `First HPI`  = round(first(Konut_Endeksi), 1),
    `Latest HPI` = round(last(Konut_Endeksi), 1),
    `Growth (%)`  = round((last(Konut_Endeksi) / first(Konut_Endeksi) - 1) * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(`Growth (%)`)) %>%
  mutate(across(c(`First HPI`, `Latest HPI`), ~ format(.x, big.mark = ","))) %>%
  kable(caption = "Table 4. HPI Growth by Region (First to Latest Observation)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

İstanbul records the highest absolute index level, while Antalya and coastal regions have also seen strong growth driven by tourism and migration. Notably, interior Anatolian regions started from a much lower base but achieved comparable percentage growth, confirming the surge is a nationwide phenomenon rather than a metropolitan one.

## Model Fitting

To formally quantify the drivers of housing price growth, we estimate a multiple linear regression model. Log-transformations are applied to all variables to handle exponential growth and to interpret coefficients as elasticities.

```{r}
#| label: model-data

model_data <- combined %>%
  mutate(
    log_HPI   = log(Konut_Endeksi),
    log_USD   = log(USD),
    log_EUR   = log(EUR),
    log_Sales = log(Total_Sales + 1),
    log_CC    = log(Total_CC + 1)
  ) %>%
  filter(!is.na(log_HPI), !is.na(log_USD))
```

### Baseline Model — Exchange Rate Only

```{r}
#| label: model1

m1 <- lm(log_HPI ~ log_USD, data = model_data)

as.data.frame(summary(m1)$coefficients) %>%
  mutate(
    Term         = c("Intercept", "log(USD/TRY)"),
    Estimate     = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    `t value`    = round(`t value`, 2),
    `p value`    = format(round(`Pr(>|t|)`, 4), nsmall = 4)
  ) %>%
  select(Term, Estimate, `Std. Error`, `t value`, `p value`) %>%
  kable(caption = sprintf("Table 5. Baseline Model – R² = %.3f",
                          summary(m1)$r.squared)) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

The exchange rate alone accounts for over 94% of the variance in the HPI. A 1% increase in the USD/TRY rate is associated with approximately a 1% increase in the housing price index.

### Full Model — Exchange Rate, Sales, and Credit Card Expenditure

```{r}
#| label: model2

m2 <- lm(log_HPI ~ log_USD + log_Sales + log_CC, data = model_data)

as.data.frame(summary(m2)$coefficients) %>%
  mutate(
    Term         = c("Intercept", "log(USD/TRY)", "log(Sales)", "log(CC Spend)"),
    Estimate     = round(Estimate, 3),
    `Std. Error` = round(`Std. Error`, 3),
    `t value`    = round(`t value`, 2),
    `p value`    = format(round(`Pr(>|t|)`, 4), nsmall = 4),
    Sig = case_when(
      `Pr(>|t|)` < 0.001 ~ "***",
      `Pr(>|t|)` < 0.01  ~ "**",
      `Pr(>|t|)` < 0.05  ~ "*",
      TRUE               ~ ""
    )
  ) %>%
  select(Term, Estimate, `Std. Error`, `t value`, `p value`, Sig) %>%
  kable(caption = sprintf("Table 6. Full Model – R² = %.3f, Adjusted R² = %.3f",
                          summary(m2)$r.squared,
                          summary(m2)$adj.r.squared)) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```

Adding sales volume and credit card expenditure raises the adjusted R² to approximately 0.95. Log(USD/TRY) remains the dominant and most significant predictor. Log(Sales) enters negatively, confirming that price increases coincide with falling transaction volumes rather than rising demand. Log(CC Spend) is positive, reflecting the general inflationary environment.

### Actual vs. Fitted

```{r}
#| label: fitted-chart

model_data$Fitted    <- fitted(m2)
model_data$Residuals <- residuals(m2)

ggplot(model_data, aes(x = Date)) +
  geom_line(aes(y = log_HPI, color = "Actual"), linewidth = 0.9) +
  geom_line(aes(y = Fitted,  color = "Fitted"),
            linewidth = 0.9, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2c7bb6", "Fitted" = "#d7191c")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Figure 10. Actual vs. Fitted log(HPI) – Full Model",
       x = NULL, y = "log(House Price Index)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

### Residual Plot

```{r}
#| label: residual-plot

ggplot(model_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.5, color = "#2c7bb6") +
  geom_hline(yintercept = 0, color = "#d7191c", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "#f46d43", linewidth = 0.8) +
  scale_x_continuous(labels = label_number(accuracy = 0.1)) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(title = "Figure 11. Residuals vs. Fitted Values",
       x = "Fitted log(HPI)", y = "Residuals") +
  theme_minimal(base_size = 11)
```

The residual plot shows that errors are small and close to zero for most of the fitted range. A slight curvature appears at the upper end (high fitted values corresponding to 2022–2023), indicating mild non-linearity during the most extreme depreciation phase — the period most difficult to model with a simple log-linear specification.

The model closely tracks the actual index throughout the sample, including the inflection point around 2021. The largest residuals cluster in 2022, coinciding with the sharpest TRY depreciation — possibly reflecting structural breaks not captured by the log-linear form.

## Limitations

No analysis is without constraints, and acknowledging them is an important part of academic work.

-   **Nominal variables only.** All variables are nominal — no inflation adjustment has been applied. Many series rise together simply because of broad inflation, which may inflate the apparent strength of correlations. A real House Price Index (deflated by CPI) would be a more precise measure of genuine affordability loss.

-   **Correlation is not causation.** The high correlations and regression fit confirm strong co-movement, but they do not establish a causal direction. Exchange rates and housing prices are likely jointly influenced by monetary policy, which we do not model explicitly.

-   **Aggregated credit card data.** The credit card expenditure variable aggregates all sectors and acts mainly as a general inflation proxy. Sector-specific data (e.g., construction materials) would be more theoretically precise.

-   **Static OLS model.** The regression does not account for autocorrelation or lagged effects. Time-series models (e.g., VAR or Error Correction Models) would better capture the dynamic transmission of exchange rate shocks into housing prices over time.

# Results and Key Takeaways

-   **Exchange rate depreciation is the primary driver.** The USD/TRY rate and the national HPI move almost in lockstep (r ≈ 0.97). A 1% rise in the exchange rate is associated with approximately a 1% rise in housing prices, reflecting the dollarised cost structure of construction materials and real estate's role as an inflation hedge.

-   **Housing sales volumes declined as prices surged.** After 2022, monthly sales softened considerably even as prices accelerated. This divergence indicates affordability-driven demand compression — buyers are being priced out rather than stimulated by the market.

-   **The price surge is nationwide, not just metropolitan.** While İstanbul leads in absolute index levels, central Anatolian regions have recorded comparable percentage growth from their lower base, suggesting the underlying inflationary forces are country-wide.

-   **Consumer spending inflation amplifies the effect.** Total credit card expenditure is strongly correlated with the HPI, as broad nominal inflation raises construction costs, labour, and land values across the board.

-   **A simple OLS model explains \~95% of HPI variance**, demonstrating that macroeconomic indicators are powerful predictors of housing price trajectories in Türkiye. More advanced time-series models (VAR, ECM) could further capture lagged dynamics and structural breaks, particularly around the 2022 currency crisis.

------------------------------------------------------------------------

*Note: Portions of the Python preprocessing code were developed with assistance from AI language model tools, as disclosed in the preprocessing section. All analytical code in R, narrative interpretation, and conclusions are the authors' own work.*

# References

Central Bank of the Republic of Türkiye (TCMB). (2024). *Electronic Data Delivery System (EVDS)*. Retrieved from https://evds2.tcmb.gov.tr/
